MCR_Algae_Data_Cleaning

### 2024-07-02 mcr_algal_data_cleaning_final.Rmd
use this cleaning script with the raw data

@@

@@

#also 99 observations that do not equal 100

@@

@@

#we want to filter out 2005 and 2006 - these were "ramp up years"
#also want to filter out rows where the sum does not = 0 
wide_alldata_filtered <- comb_wide_quad_sum %>%
  dplyr::filter(Year != 2005 & Year != 2006) %>%
  dplyr::filter(row_sums == 100)

write.csv(wide_alldata_filtered, "allbenthicdata_cleaned.csv", row.names = FALSE)

@@

>>>>

data %>% 
  filter(Taxonomy_Substrate_Functional_Group == "No data")

@@

#21000 obs of 91 variables

@@

#also 99 observations that do not equal 100, which matches what we found in the long data

@@

```{r dealing with the 99s in the wide dataset}
#LE talked to Hillary and we found that of the 85 rows that have sums that do not = 100, 49 of them are because a "-1" was accidentally added to the no data column, even though data was taken. (100 percent cover - 1 = 99). This happened only in 2017. Hillary will fix this eventually on EDI, but for now I am going to manually remove these -1 from the data set so we can continue with our analyses

#show this error in the data

no_data_mistakes <- mistakes_wide_quad_level %>%
  filter(row_sums == 99)

unique(no_data_mistakes$`No data`)

#remove the -1 from the No data column

comb_wide_quad_sum$`No data`[comb_wide_quad_sum$row_sums == 99 & comb_wide_quad_sum$`No data` == -1] <- 0

#check to see if it worked

cleaned_sumrows99 <- comb_wide_quad_sum %>%
  filter(row_sums == 99)

unique(cleaned_sumrows99$`No data`)

#yay it works

#of our 85 quads that do not equal 100, 49 have been accounted for. 23 of them are from 2020, either Backreef or Fringing, which we know were averaged. This leaves only 13 quads with questions remaining.

```{r make new row sums column}
#need to update row sum now that fixes have been made
comb_wide_quad_sum_final <- comb_wide_quad_sum %>%
  dplyr::rename(row_sum_old = row_sums) %>%
  mutate(row_sums = rowSums(dplyr::select(., 8:91))) %>%
  select(-row_sum_old)

#confirm 99 is gone 
unique(comb_wide_quad_sum_final$row_sums)

#whooo

@@

#we want to filter out 2005 and 2006 - these were "ramp up years"
#also want to filter out rows where the sum does not = 0 
wide_alldata_filtered <- comb_wide_quad_sum_final %>%
  dplyr::filter(Year != 2005 & Year != 2006) %>%
  dplyr::filter(row_sums == 100)

#should be 19750 obs

#writing one file so column names will work with Noam's code
write.csv(wide_alldata_filtered, "allbenthicdata_cleaned.csv", row.names = FALSE)

#clean names so column names are not so hard to work with

wide_alldata_filtered_cleaned_column_names <- 
  wide_alldata_filtered %>%
  janitor::clean_names()

write.csv(wide_alldata_filtered_cleaned_column_names, "allbenthicdata_cleaned_colnames.csv", row.names = FALSE)

@@

#clean names so column names are not so hard to work with

wide_algaeonly_cleaned_column_names <- 
  wide_algaeonly_filtered %>%
  janitor::clean_names()

write.csv(wide_algaeonly_cleaned_column_names, "algaeonly_cleaned_colnames.csv", row.names = FALSE)
